---
title: "Getting data into R"
output:
  html_document:
    toc: yes
    toc_float: true
    toc_depth: 2
---

# Introduction

One of the realities of business analytics or data science work is that
your data could be in any number of places or formats. Examples include:

* flat files delimited with commas, tabs, or some other character
* non-flat structured text files such as JSON or XML
* text files resulting from printer friendly reports redirected to a file
* Excel files (xlsx or xls)
* relational or non-relational databases
* from the web
    - downloadable files in some format such as csv, JSON or XML
    - inside html tables or other parts of web pages
    - via APIs
* inside pdfs
* Hadoop clusters
* some other bizarre system or bizarre format :)

Back in our Intro to R session we saw how to read csv, Excel and even
a SQLite database into an R data frame. Now we'll revisit this topic and
see a few more alternatives.

## Reading CSV files

As we saw a few weeks ago, `read.csv` and `read.table` provide a way to
easily read csv files into an R data frame. The file might be local or
sitting on a webbserver somewhere.

```{r}
tomatoURL <- "https://www.jaredlander.com/data/TomatoFirst.csv"
tomatoLocal <- "data/TomatoFirst.csv"
```

R has a function called `read.table` for reading tabular text data files. 
It also has a `read.csv` function which is just a wrapper for
`read.table` with some of the arguments preset. Let's try both.

```{r}
tomato1 <- read.table(file=tomatoLocal, header=TRUE, sep=",")
tomato2 <- read.csv(tomatoLocal)
```

Use `str` to check out their structure. You'll see they are identical.

You might also notice that the character columns got automatically
converted to factors. In this case, that's what we want. But sometimes
we might have a character field that isn't a factor and might have
a jillion unique values. If the data file is big, this conversion process
can really slow stuff down. In that case it would be better to read things
in as characters and then do conversion to factors only for those
columns that we really want to be factors. 

The auto conversion of character to factor occurs because the read functions
have a parameter called `stringsAsFactors` that defaults to TRUE. 

```{r}
tomato3 <- read.csv(tomatoLocal, stringsAsFactors=FALSE)
str(tomato3)
class(tomato3$Source)
```

The auto conversion of strings to factors as well as the need for speed has
inspired alternatives to `read.table`. 

* The **readr** package has `read_delim` for reading delimited text files,
* The **data.table** package has the `fread` function.

Both are really fast and neither autoconverts strings to factors.

The **readr** package is covered in Ch 8 of [r4ds](http://r4ds.had.co.nz/). We
will just hit a few highlights here. However, Ch 8 goes into considerable
detail about how **readr** parses text files and tries to figure out the
data type of each column - frequently a non-trivial task. A few common, but tricky,
complications that arise when reading text data include:

* international numeric formats that use comma for the decimal point,
* a myriad of date and time formats,
* columns containing a mixture of numeric and text data,
* different string encoding methods; ASCII is not universal.

We'll revisit these later in the term. For now, let's keep it simple.

```{r}
library(readr)
```

```{r}
tomato4 <- read_delim(file = tomatoLocal, delim = ',')
tomato4
class(tomato4)
```

Notice that `tomato4` is a `tibble`, not a `data.frame`. Tibbles are part of the
**tidyverse** and provide a variant of data frames that:

* print metadata such as column data types when displayed,
* intelligently print as many columns as will fit on the screen,
* don't do datatype conversions such as strings to factors automatically.
* don't automatically create row names or modify column names

Tibbles are covered in Ch 7 of [r4ds](http://r4ds.had.co.nz/). You really
can think of tibbles as data frames in most cases. For some older R packages
that might not work with tibbles, you can use the `as.data.frame` function
to convert before using. The `as_tibble` function will convert a data frame
into a tibble. While there are a few more differences between tibbles
and data frames, for this class, we can use them pretty much interchangeably.

Functions in **readr** for common special cases of comma, semicolon and tab 
delimited are `read_csv`, `read_csv2`, and `read_tsv`.

See Sections 6.1-6.3 of RforE as well as the **readcsv_notes.Rmd** file from
our Intro to R session for more on reading text files, Excel files, and
databases into data frames and/or tibbles.

Section 6.4 in RforE discusses the packages **foreign** and **haven** for
reading data from common statistical packages such as SAS, SPSS and Stata.

## R binary files

R provides a few ways to save one or more objects as binary files that
can be reloaded at any time.

### RData files

`RData` files are binary files capable of handling one or more R objects of
any kind.

Let's see what objects we have in the workspace.

```{r}
ls()
```

Let's save the `tomato4` data frame (actually a tibble) and the tomatoURL (a
string).

```{r}
save(tomato4, tomatoURL, file="data/tomato.rdata")
```

Now let's remove `tomato4` from the workspace and then reload it from the
`RData` file.

```{r}
rm(tomato4) # Remove it
head(tomato4) # Prove that it's gone - this should throw an error
```

Reload it.

```{r}
load("data/tomato.rdata")
head(tomato4)
```

A few things to note:

* the `load` command reloaded ALL of the objects in the `RData` file.
* there is no way to use `load` to load just one of the objects in an `RData` file.
* when we reloaded the tomato.rdata file, we overwrote any objects in the workspace
that had the same name as an object in the file.
* there's no way to load objects from an `RData` file into an object with a new name.

## Binary RDS files

An alternative to the all or nothing nature of `RData` files are `RDS` files. These
are designed to save a single object WITHOUT its name.

```{r}
saveRDS(tomato4, "data/tomato4.rds")
```

When we read an `RDS` file we assign it to a new object.

```{r}
tomato5 <- readRDS("data/tomato4.rds")
```

The two objects, `tomato4` and `tomato5` should be identical.

```{r}
all(tomato4 == tomato5, na.rm = TRUE)
```

## Built in data and packaged data

One really nice feature of R is that it contains a number of built in datasets.
This makes it easy to get going with trying to learn R. Also, it's easy to 
compare your results with those of others who use one of the built in datasets
to show how to accomplish some task in R. In addition to R itself, many
packages come with datasets built in. We've already seen how **ggplot** incudes
the `diamonds` dataset. Loading built in data is easy with the `data` function.

```{r}
data(diamonds, package = "ggplot2")
head(diamonds)
```

Here's information on all the datasets built into the R `datasets` package.

```{r}
data()
```

## Extracting data from web sites and JSON files

Sections 6.7 and 6.8 in RforE cover this topic. We will hold off on it for now
but will return to it when we cover getting data from the web using Python.
Python has some very widely used tools for web scraping such as **BeautifulSoup**
and we'll also see that Python dictionaries have a lot in common with JSON
files. So, we'll come back to this in a few weeks.

